The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-23 paper/paper.Rmd
Modify flowchart.

Further, \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around $350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value, transformed to log scale. The plot implies that the skewness is slightly negative. We also see that the three features are overlapped each other. It indicates that some ID's minimum wages are higher than some ID's maximum wages.

@@

```{r flow-chart, fig.cap="The stages of data filtering from the raw data to get three datasets contained yowie, n means the number of ID, while n\\_obs means the number of observations.", out.width="93%"}
```

>>>>

Further, \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around $350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fig:fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value. We still can see some extreme values in maximum wages, but consider it as a natural variability of the data. In \ref{fig:fixed-feature-plot} C, we can see that after imputing the extreme value in ID=39, we can see how the the wages change over the years more clearly.

@@

```{r flow-chart, fig.cap="The stages of data cleaning from the raw data to get three datasets contained yowie. n individuals means the number of respondents included in each stage, while n observations means the number of rows in the data. The color represented the stage of data cleaning in statistical value chain (van der Loo \\& de Jonge, 2018). Pink, blue, and green represents the raw, input, and valid data, respectively.", out.width="93%"}
```

### 2021-09-23 paper/paper.tex
Modify flowchart.

Our process of cleaning builds heavily on the tidyverse approach (Wickham et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). The resulting data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

{[}REMOVE{]} Unfortunately, the \texttt{new\_data\_qnames} did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used \texttt{tidyr} (Wickham 2020b), \texttt{dplyr} (Wickham et al. 2020), and \texttt{stringr} (Wickham 2019) to do this job.

{\centering \includegraphics[width=468px]{figures/untidy-data-1}

\caption{The missing value structure of the NLSY79 raw data.}\label{fig:untidy-data}
\end{figure}

@@

\hypertarget{initial-data-analysis}{%
\subsection{Initial Data Analysis}\label{initial-data-analysis}}

According to Huebner Marianne, Vach, and Cessie (2016), Initial Data Analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis. Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

@@

Further, \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around \$350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value, transformed to log scale. The plot implies that the skewness is slightly negative. We also see that the three features are overlapped each other. It indicates that some ID's minimum wages are higher than some ID's maximum wages.

@@

In this section, we have demonstrate a workflow from getting the data from the NSLY79 database until a ready-analysed data. Referring to Loo (2018), this section shows part of statistical value chain, which is data cleaning from raw, input, and valid data. The raw demographic and employment data from the database are not in a tidy format. Hence, we perform data wrangling to get the input data. We inspect the input data to find eligible observations to the target data. One of the target data is wages data of respondents who completed up to 12th grade and participated at least in 5 rounds of survey. This needs us to join the extracted demographic and employment data.

@@

\caption{The stages of data filtering from the raw data to get three datasets contained yowie, n means the number of ID, while n\_obs means the number of observations.}\label{fig:flow-chart}
\end{figure}

@@

There are two aspects of the original data that make a direct comparison difficult. The time variable provided is ``experience in the workforce.'' It is not clear how this is calculated. One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection. We have not been able to find this information in the database. Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.

@@

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. {``Robustlmm: An r Package for Robust Estimation of Linear Mixed-Effects Models.''} \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
Loo, Mark van der. 2018. \emph{Statistical Data Cleaning with Applications in r}.

\leavevmode\hypertarget{ref-rsvg}{}%
Ooms, Jeroen. 2020. \emph{Rsvg: Render SVG Images into PDF, PNG, PostScript, or Bitmap Arrays}. \url{https://CRAN.R-project.org/package=rsvg}.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.

>>>>

Our process of cleaning builds heavily on the tidyverse approach (H. Wickham, Averick, et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). The resulting data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

{[}REMOVE{]} Unfortunately, the \texttt{new\_data\_qnames} did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used \texttt{tidyr} (H. Wickham 2020b), \texttt{dplyr} (H. Wickham, François, et al. 2020), and \texttt{stringr} (H. Wickham 2019) to do this job.

\begin{verbatim}
#> Rows: 12,686
#> Columns: 11
#> $ CASEID_1979 <int> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1~
#> $ HRP1_1979   <int> 328, 385, 365, NA, 310, NA, NA, NA, 214, NA, 337~
#> $ HRP2_1979   <int> NA, NA, NA, NA, 375, NA, NA, NA, NA, NA, 300, NA~
#> $ HRP3_1979   <int> NA, NA, 275, NA, NA, NA, NA, NA, NA, NA, NA, NA,~
#> $ HRP4_1979   <int> NA, NA, NA, NA, NA, 250, NA, NA, NA, NA, NA, NA,~
#> $ HRP5_1979   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
#> $ HRP1_1980   <int> NA, 457, 397, NA, 333, 275, 300, 394, 200, 318, ~
#> $ HRP2_1980   <int> NA, NA, 367, NA, NA, NA, NA, NA, NA, NA, NA, NA,~
#> $ HRP3_1980   <int> NA, NA, 380, NA, NA, NA, 290, NA, NA, NA, NA, NA~
#> $ HRP4_1980   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
#> $ HRP5_1980   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
\end{verbatim}

@@

\hypertarget{initial-data-analysis}{%
\subsection{Initial data analysis}\label{initial-data-analysis}}

According to Huebner, Vach, and Cessie (2016), Initial Data Analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis. Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

@@

Further, \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around \$350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fig:fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value. We still can see some extreme values in maximum wages, but consider it as a natural variability of the data. In \ref{fig:fixed-feature-plot} C, we can see that after imputing the extreme value in ID=39, we can see how the the wages change over the years more clearly.

@@

In this section, we have demonstrate a workflow from getting the data from the NSLY79 database until a ready-analysed data. Referring to Loo and Jonge (2018), this section shows part of statistical value chain, which is data cleaning from raw, input, and valid data. The raw demographic and employment data from the database are not in a tidy format. Hence, we perform data wrangling to get the input data. We inspect the input data to find eligible observations to the target data. One of the target data is wages data of respondents who completed up to 12th grade and participated at least in 5 rounds of survey. This needs us to join the extracted demographic and employment data.

@@

\caption{The stages of data cleaning from the raw data to get three datasets contained yowie. n individuals means the number of respondents included in each stage, while n observations means the number of rows in the data. The color represented the stage of data cleaning in statistical value chain (van der Loo \& de Jonge, 2018). Pink, blue, and green represents the raw, input, and valid data, respectively.}\label{fig:flow-chart}
\end{figure}

@@

There are two aspects of the original data that make a direct comparison difficult. The time variable provided is ``experience in the workforce''. It is not clear how this is calculated. One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection. We have not been able to find this information in the database. Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.

@@

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. ``Robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
Loo, Mark van der, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.

\leavevmode\hypertarget{ref-rsvg}{}%
Ooms, Jeroen. 2020. \emph{Rsvg: Render Svg Images into Pdf, Png, Postscript, or Bitmap Arrays}. \url{https://CRAN.R-project.org/package=rsvg}.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
